---
title: "R Package: allin1"
format:
  html:
    code-tools: true
toc: true
toc-depth: 2
---

*R package built for ADC 405. [Package website](https://adc-405-s26.github.io/allin1/) · [GitHub repo](https://github.com/ADC-405-S26/allin1)*

## Overview

`allin1` is an R package that bundles shortcuts for common but tedious
data-wrangling tasks, the kind of small multi-step operations you end up
rewriting constantly. Each function collapses a familiar `dplyr` pattern into a
single call. The package currently exports three functions:

- **`sort_mutate()`** — create a new column and sort the data frame by it in one step.
- **`drop_na_rows()`** — remove every row containing a missing value.
- **`top_n_percent()`** — keep the top *N* percent of rows by a chosen variable.

The package is fully documented (with a vignette and per-function reference) on its
[pkgdown website](https://adc-405-s26.github.io/allin1/); this page is a quick tour.

## Installation

```{r}
#| eval: false
# install.packages("devtools")
devtools::install_github("ADC-405-S26/allin1")
```

```{r}
#| message: false
#| warning: false
library(allin1)
```

## Usage

The examples below use `employees`, a small demo dataset that ships with the package.

### `sort_mutate()`

Add an `earnings` column (`hours_worked * hourly_wage`) and sort by it, highest first — a mutate and an arrange in a single call:

```{r}
employees |>
  sort_mutate("earnings", hours_worked * hourly_wage, desc = TRUE)
```

### `drop_na_rows()`

Drop any row with a missing value — handy for quick cleanup before an analysis:

```{r}
employees |>
  drop_na_rows()
```

### `top_n_percent()`

Keep the highest-paid 50% of employees by `hourly_wage`:

```{r}
employees |>
  top_n_percent(hourly_wage, percent = 50)
```

## Documentation & source

The complete function reference, argument details, and a longer vignette live on the
package website. The full source, including the function definitions in `R/`, unit
tests in `tests/`, and the documentation, is on GitHub.

- Package website: [adc-405-s26.github.io/allin1](https://adc-405-s26.github.io/allin1/)
- Source code: [github.com/ADC-405-S26/allin1](https://github.com/ADC-405-S26/allin1)
