---
title: "SQL Project: Competitive Pokémon Database"
format:
  html:
    df-print: paged
    code-fold: true
    code-tools: true
toc: true
toc-depth: 2
---

*Team project with Martin Tran for STAT 405, Advanced Data Computing. Full code and data on [GitHub](https://github.com/ADC-405-S26/Project_2_SQL_Team_3).*

## Overview

Using SQL (run through a local **duckdb** database in R), we built a relational
database from three competitive-Pokémon datasets — `pokemon`, `moves`, and
`pokemon_info` — covering every Pokémon up to Generation 8, then queried it to
answer questions about competitive tiers, typings, base stats, and move power.

One result we liked: even though *Normal* has the single highest-powered move in
the game, *Fire* moves have the highest **median** power — Fire is consistently
strong, while Normal's average is dragged down by many weak moves.

Code on this page is collapsed by default. Click **► Code** on any block to
see the SQL or R behind a result. 

## Database setup

Rather than letting duckdb infer the schema automatically, we define each table
explicitly — typing every field, setting defaults, and assigning a primary key —
then load the CSVs with `COPY`. This makes the structure of the data deliberate
and reproducible.

The full construction is in the dropdown below.

::: {.callout-note collapse="true"}
## Show database setup code (schemas + loading)

```{r}
#| message: false
#| warning: false
#| code-fold: false

library(tidyverse)
library(duckdb)    # build a duckdb database locally

con_pokemon <- DBI::dbConnect(duckdb::duckdb(), dbdir = "pokemon_db")
```

```{sql connection=con_pokemon}
--| code-fold: false
USE pokemon_db;
```

```{sql connection=con_pokemon}
--| code-fold: false
DROP TABLE IF EXISTS pokemon;
```

```{sql connection=con_pokemon}
--| code-fold: false
CREATE TABLE pokemon (
  name       VARCHAR(255) NOT NULL DEFAULT ' ',  -- competitive name; unique, so primary key
  hp         SMALLINT NOT NULL DEFAULT 0,         -- base stats top out ~255, SMALLINT is plenty
  atk        SMALLINT NOT NULL DEFAULT 0,
  def        SMALLINT NOT NULL DEFAULT 0,
  spa        SMALLINT NOT NULL DEFAULT 0,
  spd        SMALLINT NOT NULL DEFAULT 0,
  spe        SMALLINT NOT NULL DEFAULT 0,
  weight     DOUBLE NOT NULL DEFAULT 0,           -- kg, has decimals -> DOUBLE
  height     DOUBLE NOT NULL DEFAULT 0,           -- m, has decimals -> DOUBLE
  formats    VARCHAR(255),                        -- competitive tier (OU, UU, NFE, LC...)
  generation VARCHAR(255),                        -- game generation code (RB, etc.)
  PRIMARY KEY (name)
);
```

```{sql connection=con_pokemon}
--| code-fold: false
COPY pokemon FROM 'pokemon_df.csv' (FORMAT CSV, HEADER, NULL 'NA');
```

```{sql connection=con_pokemon}
--| code-fold: false
DROP TABLE IF EXISTS moves;
```

```{sql connection=con_pokemon}
--| code-fold: false
CREATE TABLE moves (
  name        VARCHAR(255) NOT NULL DEFAULT ' ',  -- move name; unique, so primary key
  category    VARCHAR(255),                        -- Physical / Special / Non-Damaging
  power       SMALLINT NOT NULL DEFAULT 0,         -- base power; 0 for status moves
  accuracy    SMALLINT NOT NULL DEFAULT 0,         -- 0-100
  priority    SMALLINT NOT NULL DEFAULT 0,         -- can be negative; SMALLINT is signed
  pp          SMALLINT NOT NULL DEFAULT 0,         -- power points
  description VARCHAR(255),                         -- free-text effect
  type        VARCHAR(255),                         -- elemental type
  PRIMARY KEY (name)
);
```

```{sql connection=con_pokemon}
--| code-fold: false
COPY moves FROM 'moves_df.csv' (FORMAT CSV, HEADER, NULL 'NA');
```

```{sql connection=con_pokemon}
--| code-fold: false
DROP TABLE IF EXISTS pokemon_info;
```

```{sql connection=con_pokemon}
--| code-fold: false
CREATE TABLE pokemon_info (
  id              INTEGER NOT NULL DEFAULT 0,        -- unique row id, so primary key
  pokemon         VARCHAR(255) NOT NULL DEFAULT ' ', -- name; links to pokemon.name
  species_id      INTEGER NOT NULL DEFAULT 0,
  height          DOUBLE NOT NULL DEFAULT 0,
  weight          DOUBLE NOT NULL DEFAULT 0,
  base_experience SMALLINT NOT NULL DEFAULT 0,
  type_1          VARCHAR(255),                      -- primary type
  type_2          VARCHAR(255) DEFAULT NULL,         -- single-type pokemon have none -> NULL
  hp              SMALLINT NOT NULL DEFAULT 0,
  attack          SMALLINT NOT NULL DEFAULT 0,
  defense         SMALLINT NOT NULL DEFAULT 0,
  special_attack  SMALLINT NOT NULL DEFAULT 0,
  special_defense SMALLINT NOT NULL DEFAULT 0,
  speed           SMALLINT NOT NULL DEFAULT 0,
  color_1         VARCHAR(255),                      -- hex colour
  color_2         VARCHAR(255) DEFAULT NULL,         -- optional, so nullable
  color_f         VARCHAR(255) DEFAULT NULL,
  egg_group_1     VARCHAR(255),
  egg_group_2     VARCHAR(255) DEFAULT NULL,         -- many have a single egg group -> NULL
  url_icon        VARCHAR(255),
  generation_id   SMALLINT DEFAULT NULL,             -- 1-8; regional forms are NULL
  url_image       VARCHAR(255),
  PRIMARY KEY (id)
);
```

```{sql connection=con_pokemon}
--| code-fold: false
COPY pokemon_info FROM 'pokemon_info_df.csv' (FORMAT CSV, HEADER, NULL 'NA');
```

A quick check that all three tables built and loaded correctly:

```{sql connection=con_pokemon}
--| code-fold: false
SHOW TABLES;
```

```{sql connection=con_pokemon}
--| code-fold: false
SELECT * FROM pokemon LIMIT 5;
```
:::

## Querying the database

With the database in place, we can answer questions that would be tedious to look
up by hand, those being filtering Pokémon by tier, typing, stats, and more. Each query below runs directly against the database.

### The heaviest Pokémon in the OU tier

```{sql connection=con_pokemon}
-- heaviest pokemon in the OU (overused) competitive format (Muizz)
SELECT name, weight, formats
FROM pokemon
WHERE formats = 'OU'
ORDER BY weight DESC
LIMIT 1;
```

*The single heaviest Pokémon among those competing in the top "OverUsed" tier.*

### Strongest Pokémon in OU, by total base stats

```{sql connection=con_pokemon}
-- base stat total of pokemon in OU tier, highest to lowest (Martin)
SELECT pokemon, base_experience, formats, (pokemon_info.hp + pokemon_info.attack + pokemon_info.defense + pokemon_info.special_attack + pokemon_info.special_defense + pokemon_info.speed) AS total
FROM pokemon_info
JOIN pokemon ON pokemon.name = pokemon_info.pokemon
WHERE formats = 'OU'
ORDER BY total DESC
LIMIT 10;
```

*Joins the two tables and sums the six base stats on the fly to rank the most
statistically powerful Pokémon in the OU tier.*

### The tallest Pokémon in each generation

```{sql connection=con_pokemon}
-- tallest pokemon in each generation (Muizz)
-- Uses a correlated subquery: for each row, compare its height to the MAX height
-- within that same generation, keeping only the matches.
SELECT generation_id, pokemon, height
FROM pokemon_info AS p1
WHERE height = (
        SELECT MAX(height)
        FROM pokemon_info AS p2
        WHERE p1.generation_id = p2.generation_id
      )
  AND generation_id IS NOT NULL;   -- regional forms have no listed generation (real NULLs)
```

*A correlated subquery returns both the Pokémon name and its generation in one
result which is something a simple `GROUP BY MAX(height)` can't do on its own.*

### Poison-type Pokémon in the UU tier

```{sql connection=con_pokemon}
-- poison types in the UU (underused) format (Muizz)
SELECT pokemon, type_1, type_2, pokemon.formats
FROM pokemon_info
JOIN pokemon ON pokemon_info.pokemon = pokemon.name
WHERE (type_1 = 'poison' OR type_2 = 'poison')
  AND pokemon.formats = 'UU';
```

*Checks both type slots so dual-typed Poison Pokémon aren't missed.*

### Monster egg-group Pokémon, by generation (NFE / LC tiers)

```{sql connection=con_pokemon}
-- count of NFE (not fully evolved) or LC (little cup) Monster egg-group pokemon
-- per generation (Muizz)
SELECT generation_id AS gen, COUNT(*) AS num_pokemon
FROM pokemon_info
JOIN pokemon ON pokemon_info.pokemon = pokemon.name
WHERE (pokemon.formats = 'NFE' OR pokemon.formats = 'LC')
  AND (egg_group_1 = 'monster' OR egg_group_2 = 'monster')
GROUP BY gen;
```

*Combines a join, multiple tier filters, and a grouped count in one query.*

### Quick stats

A few smaller lookups against the same database.

How many moves have non-neutral priority?

```{sql connection=con_pokemon}
-- number of moves with priority other than 0 (Martin)
SELECT COUNT(*) AS num_rows
FROM moves
WHERE priority <> 0;
```

The ten lightest Pokémon:

```{sql connection=con_pokemon}
-- top 10 lightest pokemon (Martin)
SELECT pokemon, weight
FROM pokemon_info
ORDER BY weight
LIMIT 10;
```

Count of Pokémon by primary type, most common first:

```{sql connection=con_pokemon}
-- number of pokemon per primary type, highest to lowest (Martin)
SELECT type_1, COUNT(*) AS type_count
FROM pokemon_info
GROUP BY type_1
ORDER BY type_count DESC;
```

## Visualizations

### Average base-stat total by generation

Which generation introduced the strongest Pokémon on average? We average the six
base stats across every Pokémon in each generation.

```{sql connection=con_pokemon}
--| code-summary: "SQL: average stat total per generation"
SELECT generation_id,
       AVG(hp + attack + defense + special_attack + special_defense + speed) AS total
FROM pokemon_info
WHERE generation_id IS NOT NULL   -- exclude regional forms with no listed generation
GROUP BY generation_id;
```

```{r}
#| code-summary: "Plot: average stat total per generation"
#| message: false
#| warning: false

generation_plot <- DBI::dbGetQuery(con_pokemon,
  "SELECT generation_id,
          AVG(hp + attack + defense + special_attack + special_defense + speed) AS total
   FROM pokemon_info
   WHERE generation_id IS NOT NULL
   GROUP BY generation_id;"
)

ggplot(generation_plot, aes(x = generation_id, y = total)) +
  geom_col(aes(fill = total)) +
  geom_text(aes(label = round(total)), nudge_y = 10) +
  scale_fill_gradient(low = "white", high = "red") +
  labs(x = "Generation", y = "Average Stat Total",
       title = "Average Base Stats by Generation") +
  theme(legend.position = "none")
```

*Average base-stat totals stay fairly close across generations, with later
generations trending slightly higher.*

### Move power distribution by type

How hard does each move type hit? We look at the spread of base power per type,
excluding status moves (which have 0 power).

```{sql connection=con_pokemon}
--| code-summary: "SQL: damaging-move power by type"
SELECT type, power
FROM moves
WHERE power > 1   -- drop status / non-damaging moves
ORDER BY power DESC;
```

```{r}
#| code-summary: "Plot: move power distribution by type"
#| message: false
#| warning: false

type_plot <- DBI::dbGetQuery(con_pokemon,"SELECT type, power FROM moves WHERE power > 1")

ggplot(type_plot, aes(x = reorder(type, power, FUN = median), y = power, fill = type)) +
  geom_boxplot(alpha = 0.6) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45), legend.position = "none") +
  labs(title = "Distribution of Move Power by Type",
       x = "Move Type", y = "Base Power")
```

*Ordered by median power: even though Normal has the single highest-powered move,
its median is near the bottom — while Fire is consistently high, with the top median.*

```{r}
#| echo: false
DBI::dbDisconnect(con_pokemon, shutdown = TRUE)
```

## Sources & tools

- Competitive Pokémon tier dataset (`pokemon_df`, `moves_df`) — [Kaggle](https://www.kaggle.com/datasets/varlawend/competitive-pokmon-tier-dataset)
- Pokémon info dataset (`pokemon_info_df`) — [TidyTuesday, 2025-04-01](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-04-01)
- AI tools (LLM) were used as an aid for dataset editing during development.
